---
title: "report 2"
author: "Victor Jacobs"
date: "2026-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r data wrangling}
library(readxl)
library(tidyverse)
library(dplyr)
library(tidyr)

peer_raw <- read_excel(
  "Short Report 02_Data (1).xlsx",
  sheet = "1_Peer Group Salaries",
  col_names = FALSE)

peer_unadj <- peer_raw %>%
  slice(5:n()) %>%  
  select(1:5)

colnames(peer_unadj) <- c("institution", "all_fac","full", "associate", "assistant")

peer_unadj_long <- peer_unadj %>%
  pivot_longer(cols = c(assistant, associate, full), names_to = "rank", values_to = "salary") %>%
  mutate(adjustment = "Unadjusted")

peer_cola <- peer_raw %>%
  slice(5:n()) %>%
  select(6:10)

colnames(peer_cola) <- c("institution", "all_fac", "full", "associate", "assistant")

peer_cola_long <- peer_cola %>%
  pivot_longer(cols = c(assistant, associate, full), names_to = "rank",
    values_to = "salary") %>%
  mutate(adjustment = "COLA-adjusted")

peer_long <- bind_rows(peer_unadj_long, peer_cola_long) %>%
  mutate(rank = factor(rank, levels = c("assistant", "associate", "full")),
    adjustment = factor(adjustment, levels = c("Unadjusted", "COLA-adjusted")))



davidson_salary <- read_excel("Short Report 02_Data (1).xlsx", sheet = "2_Inf-Adj Davidson Salaries")

colnames(davidson_salary) <- c("year", "all", "full", "associate", "assistant")

davidson_long <- davidson_salary %>% pivot_longer(-year, names_to = "rank", values_to = "salary_real")



tuition <- read_excel("Short Report 02_Data (1).xlsx", sheet = "3_Inf_Adj Tuition Fees")

colnames(tuition) <- c("year", "tuition_nominal", "inflation_factor", "tuition_real")

tuition <- tuition %>% select(year, tuition_real)



endowment <- read_excel("Short Report 02_Data (1).xlsx", sheet = "5_Inf_Adj Endowment")

colnames(endowment) <- c("year", "endowment_real")



endow_detail <- read_excel("Short Report 02_Data (1).xlsx", sheet = "6_Inf_Adj Endow Faculty Student")



expenses <- read_excel("Short Report 02_Data (1).xlsx", sheet = "7_Davidson Expenses")

colnames(expenses) <- c("category", "amount")

expenses <- expenses %>%
  filter(category != "Total expenses (F1920)")
```



```{r 1. strip plot}

library(ggplot2)


ggplot(peer_long, aes(x = rank, y = salary, color = rank)) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2, show.legend = FALSE) +
  stat_summary(fun = mean, geom = "point", color = "black", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", color = "black",
    width = 0.15) +
  facet_wrap(~ adjustment) +
  labs(title = "Faculty Salaries at Peer Institutions by Rank",
    subtitle = "Black points and bars show mean ± standard error",
    x = "Faculty Rank",y = "Salary") +
  theme_minimal()

```

Faculty salaries at Davidson are consistently lower than those at peer institutions across all ranks.

```{r 2. Line plot}

ggplot(davidson_long, aes(x = year, y = salary_real, color = rank)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Inflation-Adjusted Faculty Salaries at Davidson College",
    subtitle = "By Rank, Across Fiscal Years", x = "Fiscal Year",
    y = "Salary (Inflation-Adjusted $)", color = "Rank") + 
  theme_minimal()


```

Inflation-adjusted faculty salaries at Davidson have grown only little over time.

```{r 3. Heatmap}

ggplot(tuition, aes(x = year, y = 1, fill = tuition_real)) +
  geom_tile(color = "white") +
   labs(title = "Inflation-Adjusted Tuition and Fees Over Time",
    subtitle = "Darker shading indicates higher tuition levels", x = "Fiscal Year", y = NULL, fill = "Tuition") +
  theme_minimal()
 
```

Inflation-adjusted tuition and fees at Davidson have steadily increased over time.


```{r 4. scatterplot}
library(dplyr)

avg_salary <- davidson_long %>%
  group_by(year) %>%
  summarize(avg_salary = mean(salary_real, na.rm = TRUE))

merged <- avg_salary %>%
  left_join(tuition, by = "year")


ggplot(merged, aes(x = tuition_real, y = avg_salary)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Faculty Salaries vs Tuition Over Time",
    subtitle = "Each point represents a fiscal year",
    x = "Tuition and Fees (Inflation-Adjusted $)",
    y = "Average Faculty Salary (Inflation-Adjusted") +
    theme_minimal()
```

Increases in tuition have outpaced growth in average faculty salaries over time.


```{r 5. Bar plot}

ggplot(endowment, aes(x = factor(year), y = endowment_real)) +
  geom_col() +
  labs(title = "Inflation-Adjusted Endowment Over Time",
    subtitle = "Davidson College Endowment by Fiscal Year", x = "Fiscal Year",
    y = "Endowment (Inflation-Adjusted)") +
  theme_minimal()

```
Davidson College’s inflation-adjusted endowment has grown substantially over time.


```{r 6. Bubble Plot}

ggplot(endow_detail, aes(x = Year...1, y = `Per Student`, size = `Per Faculty Member`)) +
  geom_point(alpha = 0.6) +
  labs(title = "Inflation-Adjusted Endowment per Student Over Time",
    subtitle = "Bubble size represents inflation-adjusted endowment per faculty member", x = "Fiscal Year", y = "Endowment per Student (Inflation-Adjusted)", size = "Endowment per Faculty") +
  theme_minimal()


```
Endowment resources per student and per faculty member have increased over time, indicating growing institutional capacity.

```{r 7. Treemap}

library(treemapify)

ggplot(expenses, aes(area = amount, fill = category, label = category)) +
  geom_treemap() +
  geom_treemap_text(colour = "white", place = "centre", reflow = TRUE) +
  labs(title = "Davidson College Expenditures", subtitle = "Latest Fiscal Year Available") +
  theme_minimal()

```

Faculty compensation represents only a portion of total institutional expenditures relative to other spending categories.



